---
title: "P1-ethanol"
author: "Aidan Burk"
date: "2024-04-28"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r metadata, echo=FALSE}
# Author:  Aidan Burk
# Date:    2024-04-28
# Purpose: Portfolio 1 - Ethanol Assignment rmd file
#-------------------------------------------------------------------------------
```

```{r packages, include=FALSE}
# Load packages
library("tidyverse")
library("knitr")
library("ggResidpanel")
```

```{r load data}
# Read & clean data
gas_mileage <- read.csv('gas_mileage_data.csv') |>
  drop_na(c(mpg, ethanol)) |> # remove rows with NA mpg/ethanol
  mutate(
    ethanol = ifelse(ethanol == 10,
                     yes = '10 %',
                     no  = '0 %') # rewrite 0/10 values in ethanol as Yes/No
  )
```

## Introduction

gas_mileage_data.csv consists of data that Professor Jarad Niemi
collected on `r nrow(gas_mileage)` trips made on his 2011 Toyota Sienna.
At each gas fill up, he recorded the number of miles on the trip meter
(then reset the trip meter) as well as the amount of gas and the type of
gas in terms of the octane and ethanol percentage. This analysis is
meant to answer the question: How is gas mileage (miles per gallon)
different when using gas with no ethanol compared to gas with 10%
ethanol?

## Data

```{r summary-statistics}
s <- gas_mileage |>
  group_by(ethanol) |>
  summarize(Obervations    = n(),
            `Mean mpg` = mean(mpg),
            `Sd mpg`   = sd(mpg),
            .groups = 'drop')
```

```{r summary-statistics-table}
# Create HTML table from summary statistics}
s |>
  rename(Ethanol = ethanol) |>
  knitr::kable(
    caption = "Summary of miles per gallon for gas with no ethanol & gas with 10% ethanol",
    # label   = "summary-statistics",
    align   = c("rlll")) # Column alignment: [r]ight or [l]eft
```

The figure below displays a scatterplot of each trip's mileage versus
type of gas (No ethanol or 10% ethanol). When comparing gas with no
ethanol to gas with 10% ethanol, this figure suggests gas with no
ethanol has slightly higher mean mpg, with a range of
`r gas_mileage |> filter(ethanol == '0 %') |> pull(mpg) |> min() |> round(2)` to
`r gas_mileage |> filter(ethanol == '0 %') |> pull(mpg) |> max() |> round(2)`, which
is wider than the range of gas with 10% ethanol
(`r gas_mileage |> filter(ethanol == '10 %') |> pull(mpg) |> min() |> round(2)` to
`r gas_mileage |> filter(ethanol == '10 %') |> pull(mpg) |> max() |> round(2)`).

```{r plot}
# Create exploratory scatterplot
ggplot(gas_mileage,
       aes(x = ethanol, 
           y = mpg)) + 
  geom_jitter(height = 0,
              width  = 0.1,
              color = '#101020',
              size = 1) +
  labs(title = 'Scatterplot - Mileage by Type of Gas',
       x = 'Type of gas (Ethanol percentage)',
       y = 'Miles per gallon (mpg)') +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5))

```

## Results

```{r model}
# Fit regression model of the logarithm of mpg on ethanol
m <- lm(log(mpg) ~ ethanol, 
         data = gas_mileage)
```

Because the explanatory variable (Type of gas) is categorical, & the
residuals have a skewed distribution, we will fit a log-linear
regression model for the logarithm of miles per gallon on type of gas.
The structure of the logarthimic regression model is
$$log(Y_i)=\beta_0+\beta_1X_i+\epsilon_i$$ where $Y_i$ is miles per
gallon, $X_i$ is the type of gas (0% = 0, 10% = 1), & $\epsilon_i$
is the error.

The 95% confidence interval for the multiplicative effect of mpg when
the gas has ethanol compared to when it does not is
`r exp(confint(m)[2,]) |> round(2)`. This suggests that with 95%
confidence, the multiplicative change in mpg from having no ethanol to
having 10% ethanol $e^{\beta_0}$ is between
`r exp(confint(m)[2,1]) |> round(2)` &
`r exp(confint(m)[2,2]) |> round(2)`.

Diagnostic plots are included below to help evaluate model assumptions.

```{r diagnostic plots}
# Create panel of diagnostic plots
resid_panel(m,
            plots = c("resid",
                      "index",
                      "qq",
                      "cookd"))
```

Below is a plot of the estimated mean miles per gallon with 95%
confidence intervals. As seen with the original data, mpg generally
decreases a slight amount when the type of gas has 10% ethanol compared
to no ethanol.

```{r predictions}
# Create data frame with unique explanatory variable values
nd <- gas_mileage |>
  select(ethanol) |>
  unique()

# Predict
p <- bind_cols(
  nd,
  predict(m,
          newdata = nd,
          se.fit = TRUE)|>
      as.data.frame() |>
      
      # Manually construct confidence intervals
      mutate(
        lwr = fit - qnorm(0.975) * se.fit,
        upr = fit + qnorm(0.975) * se.fit,
        
        # Exponentiate to get to response scale
        mpg = exp(fit),
        lwr    = exp(lwr),
        upr    = exp(upr)
      ) 
  )

# Plot group estimates
ggplot(p,
       aes(x = ethanol,
           y = mpg,
           group = 1,
           ymin = lwr,
           ymax = upr)) + 
  geom_pointrange(color = '#101020') +
  geom_text(data = p,
            mapping = aes(x = ethanol, label = round(mpg, 2), y = mpg),
            vjust = -.8, hjust = -.2, size = 8/.pt) +
  geom_line() +
  labs(title = 'Estimated Mean Mileage (mpg) with 95% Confidence Intervals',
       subtitle = 'Log-linear Regression Model',
       x = 'Type of gas (Ethanol percentage)',
       y = 'Miles per gallon (mpg)') +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5))

```

## Discussion

Using data obtained from car trips made on Professor Niemi's 2011 Toyota
Sienna, recording number of miles and type of gas at each gas fill up,
we performed simple linear regression analysis, modelling the logarithm
of miles per gallon on type of gas. We found that when comparing the two
types of gas, gas with 10% ethanol is associated with a generally lower
mileage than for gas with no ethanol.

Since the data were recorded in temporal order, there is no chance for
our assumption of independence to be violated, and our diagnostic plots
show no change in residuals due to time. The normality assumption is
reasonably met as indicated by a normal quantile-quantile plot. Our
assumption of constant variance is also reasonably met as the residuals
show little to no variability.

Because the type of gas selected at each gas fill up was not randomized, our results on the effect of type of gas on mileage cannot be interpreted as causal. Since the trips made on Professor Niemi's 2011 Toyota
Sienna are not a random sample from any population, we cannot justifiably extend these results to a larger population.

