---
title: "P2-analysis.Rmd"
author: "Aidan Burk"
date: "2024-05-06"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE, purl=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```

```{r metadata, echo=FALSE}
# Author:  Aidan Burk
# Date:    2024-05-06
# Purpose: Portfolio 2 - Analysis Assignment rmd file
#-------------------------------------------------------------------------------
```

```{r packages, include=FALSE}
# Load packages
library("tidyverse")
library("knitr")
library("ggResidpanel")
library("MASS")
```

```{r load data}
# Read data
birthwt <- MASS::birthwt |>
  dplyr::select(bwt, lwt, smoke) |>
  mutate(
    smoke = ifelse(smoke == 1,
                     yes = 'Yes',
                     no  = 'No')
  )
row.names(birthwt) <- NULL
```

## Introduction

The birthwt dataset has data collected by David W. Hosmer & Stanley
Lemeshow at Baystate Medical Center, Springfield, Massachusetts during
1986, with `r nrow(birthwt)` observations, and variables including infant birth
weight in grams, mother's weight in pounds at last menstrual period, and
smoking status during pregnancy. This analysis is for studying how
mother's weight at last menstrual period and smoking status affect infant birth
weight. (Source: Hosmer, D.W. and Lemeshow, S. (1989) Applied Logistic
Regression. New York: Wiley)

## Data

```{r summary-statistics}
s <- apply(birthwt[c(1,2)],2,function(x) summary(x)) |> data.frame() |>
  t() |> as.data.frame()
s$Variable <- c('Birth weight', 'Last menstrual period weight')
s <- s[c(7,1:6)]
row.names(s) <- NULL
```

```{r numeric-summary-statistics-table}
# Create HTML table from summary statistics}
s |>
  knitr::kable(
    caption = "Summary of numeric variables")
```

```{r smoke-summary-statistics}
s1 <- birthwt |>
  group_by(smoke) |>
  summarize(Obervations    = n(),
            `Mean bwt` = mean(bwt),
            `Sd bwt`   = sd(bwt),
            .groups = 'drop')
```

```{r smoke-summary-statistics-table}
# Create HTML table from summary statistics}
s1 |>
  rename(Smoke = smoke) |>
  knitr::kable(
    caption = "Summary of birth weight for mothers who didn't smoke during pregnancy & mothers who smoked during pregnancy",
    # label   = "summary-statistics",
    align   = c("rlll")) # Column alignment: [r]ight or [l]eft
```

The figure below displays a scatterplot of mother's weight at last
menstrual period versus birth weight and smoking status. The plot
suggests a slight increase in birth weight as last menstrual period
weight increases. When comparing those who smoked during pregnancy to
those who didn't, this figure shows birth weight to generally be
slightly lower for those who smoked, with a range of
`r birthwt |> filter(smoke == 'Yes') |> pull(bwt) |> min() |> round(2)`
to
`r birthwt |> filter(smoke == 'Yes') |> pull(bwt) |> max() |> round(2)`,
which is lower and less wide compared to the range of those who didn't
smoke
(`r birthwt |> filter(smoke == 'No') |> pull(bwt) |> min() |> round(2)`
to
`r birthwt |> filter(smoke == 'No') |> pull(bwt) |> max() |> round(2)`).

```{r plot}
# Create exploratory scatterplot
g <- ggplot(birthwt,
            aes(x = lwt, 
            y = bwt,
            color = smoke)) + 
  geom_point() +
  labs(title = 'Scatterplot - Birth Weight by Last Menstrual Period Weight\n & Smoking Status During Pregnancy',
       x = 'Mother’s weight at last menstrual period (pounds)',
       y = 'Birth Weight (g)',
       color = 'Smoking status') +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5)) +
  scale_color_manual(values=c("plum1", "plum4"))

g
```

## Results

```{r model}
# Fit regression model of the logarithm of mpg on ethanol
m <- lm(bwt ~ lwt + smoke, 
         data = birthwt)
```

Because the response variable (Birth weight) is continuous and the
explanatory variables include a continuous & categorical variable, we
will fit a multiple linear regression model for birth weight on last menstrual
period weight and smoking status. A scatterplot with regression lines
indicates no significant interaction between the explanatory variables.

```{r lmplot}
# Obtain unique values of Diet
nd <- expand.grid(
  smoke = unique(birthwt$smoke),
  lwt = range(birthwt$lwt)
) 

# Construct obtain predictions
p <- nd |>
  mutate(
    bwt = predict(m, newdata = nd)
  )

# Plot
g +
  geom_line(data = p) 

```

The structure of the multiple regression model is
$$Y_i=\beta_0+\beta_1X_{i,1}+\beta_2X_{i,2}+\epsilon_i$$ where $Y_i$ is
birth weight, $X_{i,1}$ is the mother's last menstrual period weight,
$X_{i,2}$ is smoking status (No = 0, Yes = 1), & $\epsilon_i$ is the
error.

The 95% confidence interval for the effect of mother's last menstrual
period weight on birth weight is `r confint(m)[2,] |> round(2)`, which
means that with 95% confidence, when last menstrual period weight
increases by 1 pound, birth weight increases between
`r confint(m)[2,1] |> round(2)` & `r confint(m)[2,2] |> round(2)` grams.

The 95% confidence interval for the effect of smoking status on birth
weight is `r confint(m)[3,] |> round(2)`, which means that with 95%
confidence, the change in birth weight between not smoking during
pregnancy & smoking during pregnancy is a decrease between
`r confint(m)[3,2] |> round(2) |> abs()` &
`r confint(m)[3,1] |> round(2) |> abs()` grams.

Diagnostic plots are included below to help evaluate model assumptions.

```{r diagnostic plots}
# Create panel of diagnostic plots
resid_panel(m,
            plots = c("resid",
                      "index",
                      "qq",
                      "cookd"))
```


Below is a plot of the estimated birth weight with 95% confidence
intervals. As seen with the original data, birth weight generally
increases a slight amount when last menstrual period weight increases,
and is slightly higher for those who didn't smoke during pregnancy than
for those who smoked.

```{r predictions}
# Predict
p <- bind_cols(
  nd,
  predict(m,
          newdata = nd,
          se.fit = TRUE)|>
      as.data.frame() |>
      
      # Manually construct confidence intervals
      mutate(
        lwr = fit - qnorm(0.975) * se.fit,
        upr = fit + qnorm(0.975) * se.fit
      ) 
  )

# Plot group estimates
ggplot(p,
       aes(x = lwt,
           y = fit,
           group = smoke,
           color = smoke,
           ymin = lwr,
           ymax = upr)) + 
  geom_pointrange() +
  geom_text(data = p,
            mapping = aes(x = lwt, label = round(fit, 2), y = fit),
            vjust = .8, hjust = -.2, size = 8/.pt) +
  geom_line() +
  labs(title = 'Estimated Mean Birth Weight (g) with 95% Confidence Intervals',
       subtitle = 'Multiple Regression Model',
       x = 'Mother’s weight at last menstrual period (pounds)',
       y = 'Birth Weight (g)',
       color = 'Smoking status') +
  theme_minimal() +
  theme(plot.title = element_text(hjust = 0.5),
        plot.subtitle = element_text(hjust = 0.5)) +
  scale_color_manual(values=c("plum1", "plum4"))

```

## Discussion

Using data collected from Baystate Medical Center, Springfield, Massachusetts by David W. Hosmer & Stanley Lemeshow in 1986, we performed multiple
linear regression analysis, modelling infant birth weight
on mother's weight at last menstrual period and smoking status during pregnancy. We found that increasing weight at last menstrual period is associated with a slight increase in infant birth weight, and when comparing smoking status, smoking during pregnancy is associated with a generally lower birth weight than not smoking during pregnancy.

We don't have enough information to assess our independence assumption, and the data does not appear to be in temporal order. The normality assumption is
reasonably met as indicated by a normal quantile-quantile plot. Our
assumption of constant variance is also reasonably met as the residuals
show little to no variability.

Because the data were not randomized as far as we know,
our results on the effect of last menstrual
period weight and smoking status on infant birth weight cannot be
interpreted as causal, and since the data are not a random sample from any population, we cannot
justifiably extend these results to a larger population.
